---
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## Project Overview and Question of Interest
The data for this project is at https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic. It lists shooting incidents that occurred in NYC from 2006 through the end of the last year. Per the site:

>"This is a breakdown of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year. This data is manually extracted every quarter and reviewed by the Office of Management Analysis and Planning before being posted on the NYPD website. Each record represents a shooting incident in NYC and includes information about the event, the location and time of occurrence. In addition, information related to suspect and victim demographics is also included. This data can be used by the public to explore the nature of shooting/criminal activity. Please refer to the attached data footnotes >for additional information about this dataset."

The first thing I'm interested in is exploring the overall counts and percentages for boroughs and precincts as tables. And that of victim's sex and race. Turning to visualization, I'm curious about how to best to present shootings by time of day; opting for a heat map. I then turn my attention to visualizing murder versus non-murder counts per year, expressed as a stacked bar chart. Finally, the model explores the probability (if shot) of getting murdered by borough.


## 1. Import libraries, data, and prepare.
```{r setup, message=FALSE}
   library(tidyverse) # A collection of data science packages.
   library(lubridate) # A package that makes it easier to work with dates and times.
   library(knitr) # For tables in PDF/HTML.
```

```{r import, message=FALSE}
   # Import the data, reading the csv in as a data frame.
   data0 = read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
```

```{r limit, message=FALSE}
   # Choose the columns I think are relevant to my analysis.
   data = data0 %>%
   select (
      OCCUR_DATE,
      OCCUR_TIME,
      BORO,
      PRECINCT,
      STATISTICAL_MURDER_FLAG,
      PERP_AGE_GROUP,
      PERP_SEX,
      PERP_RACE,
      VIC_AGE_GROUP,
      VIC_SEX,
      VIC_RACE
   )
```

```{r replace, message=FALSE}
   # Replace empty values in the data set. Unsurprisingly, the perpetrator data is spotty.
   data = data %>% 
   replace_na(list(PERP_AGE_GROUP = "UNKNOWN", PERP_SEX = "UNKNOWN", PERP_RACE = "UNKNOWN"))
```


\newpage
## 2. Analyze
```{r analyze-borough, message=FALSE}
   # Counts and percentages of shooting incidents by borough.
   data %>%
   count(BORO) %>%
   mutate(Percent = round(n / sum(n) * 100, 1)) %>%
   arrange(desc(n)) %>%
   kable(caption = "Counts and Percentages by Borough")
```

\vspace{2em}
```{r analyze-precinct, message=FALSE}
   # Top 10 precincts with the most shootings. Sorted descending.
   data %>%
   count(PRECINCT, sort = TRUE) %>%
   head(10) %>%
   kable(caption = "Top 10 Precincts with the Most Shootings")
```

\newpage
```{r analyze-sex, message=FALSE}
   # Victim sex distribution.
   data %>%
   count(VIC_SEX) %>%
   mutate(Percent = round(n / sum(n) * 100, 1)) %>%
   kable(caption = "Victim Sex Distribution")
```

\vspace{2em}
```{r analyze-race, message=FALSE}
   # Victim race distribution.
   data %>%
   count(VIC_RACE) %>%
   mutate(Percent = round(n / sum(n) * 100, 1)) %>%
   arrange(desc(n)) %>%
   kable(caption = "Victim Race Distribution")
```

\newpage
## 3. Visualize
```{r visualize-heatmap, message=FALSE}
   # Visualize incidents by day of week and time of day.
   hour_levels <- format(strptime(0:23, format = "%H"), "%I %p")
   hour_levels <- factor(hour_levels, levels = rev(hour_levels))  # Midnight at top

   # Prepare data
   heat_data <- data %>%
   mutate(
   OCCUR_DATE = as.Date(OCCUR_DATE, format = "%m/%d/%Y"),
   HOUR = hour(hms(OCCUR_TIME)),
   HOUR_LABEL = factor(format(strptime(HOUR, format = "%H"), "%I %p"), levels = levels(hour_levels)),
   DOW = wday(OCCUR_DATE, label = TRUE, abbr = FALSE)
   ) %>%
   count(DOW, HOUR_LABEL) 

   # Plot heatmap
   ggplot(heat_data, aes(x = DOW, y = HOUR_LABEL, fill = n)) +
   geom_tile(color = "white", linewidth = 0.3) +
   scale_fill_gradient(low = "#F1F1F1", high = "brown", name = "Incidents") +
   labs(
   title = "NYC Shootings by Time of Day",
   x = "Day of Week",
   y = "Hour"
   ) + theme_minimal(base_family = "serif")
```

\newpage
```{r visualize-barchart, message=FALSE}
   # Count of shooting incidents per year and outcome.

   # Prepare data
   stacked_data <- data %>%
     mutate(
       OCCUR_DATE = as.Date(OCCUR_DATE, format = "%m/%d/%Y"),
       YEAR = year(OCCUR_DATE),
       MURDER_FLAG = case_when(
         STATISTICAL_MURDER_FLAG == TRUE ~ "Murder",
         STATISTICAL_MURDER_FLAG == FALSE ~ "Non-Murder",
         TRUE ~ NA_character_
       )
     ) %>%
     filter(!is.na(YEAR), YEAR >= 2006 & YEAR <= 2024, !is.na(MURDER_FLAG)) %>%
     mutate(MURDER_FLAG = factor(MURDER_FLAG, levels = c("Non-Murder", "Murder"))) %>%
     count(YEAR, MURDER_FLAG)

   # Plot with murders stacked at bottom
   ggplot(stacked_data, aes(x = as.factor(YEAR), y = n, fill = MURDER_FLAG)) +
     geom_bar(stat = "identity") +
     labs(
       title = "NYC Shooting Incidents by Year and Outcome (2020–2024)",
       x = "Year",
       y = "Number of Incidents",
       fill = "Incident Type"
     ) +
     scale_fill_manual(
       values = c("Murder" = "brown", "Non-Murder" = "#f2c6b4"),
       breaks = c("Murder", "Non-Murder")  # Legend order: Murder first
     ) +
     theme_minimal(base_size = 11) +
     theme(
       legend.position = "bottom",
       legend.text = element_text(size = 9),
       legend.title = element_text(size = 10),
       legend.key.size = unit(0.4, "cm")
     )
```

\newpage
## 4. Model
```{r model, warning=FALSE, message=FALSE}
   # Model and plot probability of murder if shot per borough

   # Prepare filter and format
   data_clean <- data %>%
     mutate(
       MURDER_FLAG = if_else(STATISTICAL_MURDER_FLAG == TRUE, 1, 0),
       BORO = as.factor(BORO)
     ) %>%
     filter(!is.na(MURDER_FLAG), !is.na(BORO))

   # Model: Predict murder likelihood by borough
   model <- glm(MURDER_FLAG ~ BORO, data = data_clean, family = binomial)
   summary(model)

   # Create new data frame for each borough
   newdata <- data.frame(BORO = levels(data_clean$BORO))

   # Predict probabilities for each borough
   newdata$predicted_prob <- predict(model, newdata = newdata, type = "response")

   # Plot predicted probabilities
   ggplot(newdata, aes(x = reorder(BORO, -predicted_prob), y = predicted_prob)) +
     geom_col(fill = "brown") +
     labs(
       title = "If Shot, Predicted Probability of Murder by Borough",
       x = "Borough",
       y = "Predicted Probability"
     ) + theme_minimal(base_size = 11)
```

\newpage
## 5. Conclusion and Possible Bias

In conclusion...a depressing data set! Exploring it, I looked at overall differences between the boroughs and precincts by count and percentages across all years. I explored the same with victims' sex and race. I visualized shootings by time of day. I compared murder to non-murder per year with a stacked bar chart. Finally, the model I chose was the probability of getting murdered (if shot) by borough.


Possible bias in the dataset:

* There are some data points missing for some categories, especially the perpetrator columns. These can skew group analysis.
* The dataset reflects reported shooting incidents. Those that are unreported or misclassified will be ommitted.
* Victims and/or witnesses may underreport due to a lack of trust in police, or some other fear.
* Police resources may not be evenly distributed across NYC.
* Neighborhoods with higher policing may have more recorded incidents. This may overrepresent minority communities, even if similar crimes happen elsewhere.

Possible bias in my analysis:

* My analysis may reflect COVID-era anomalies.
* In my heatmap, grouping by hour/day is useful but may not consider other patterns like season. I'm considering splitting these into small multiples per year.
* In my heatmap, readers may overinterpret minor gradients of color as more meaningful than it is.
* If I don't control for population density (for example, by normalizing per capita), I may just be highlighting more populous areas.
